---
title: "AN588-Week-2"
author: "Samantha Vee"
date: "2023-09-11"
output: html_document
---

## Preliminaries 1
```{r prelim 1}
setwd("/Users/samanthavee/Documents/Github/AN588-Week-2") #set wd
library("stringr") #loads package
```

### CHALLENGES

## Challenge 1
```{r challenge 1}
v1 <- c('Stately', ',', 'plump', 'Buck', 'Mulligan', 'came', 'from', 'the', 'stairhead', ',', 'bearing', 'a', 'bowl', 'of', 'lather', 'on', 'which', 'a', 'mirror', 'and', 'a', 'razor', 'lay', 'crossed', '.'); v1
#one way to split up character 

v2 <- "Stately, plump Buck Mulligan came from the stairhead, bearing a bowl of lather on which a mirror and a razor lay crossed."; v2 #entire sentence is one character 
v2<- gsub("[[:punct:]]","",v2); v2 #remove punctuation
v2 <- str_split(v2, " "); v2 #splits up sentence so each word is its own character
v2.1 <- v2[seq(from = 1, to = length(v2), by = 3)]; v2.1 #this doesn't work because seq doesn't like characters, try again !! 

v2.1 <- str_split(v2, " ") #ignore this sam
#trying in class below
#line 18 then remove punctuation
str_split(v1, "Buck", n=1, simplify = FALSE)
#or use extract function to extract every third
```

## Challenge 2
```{r challenge 2}
m <- matrix(data = 1:80, nrow =8, ncol = 10, byrow = FALSE); m
x <- m[4,5]; x #extract element in row 4, column 5 and assign to new variable x

#extract 2nd, 3rd, 6th COLUMNS and assign to new variable x
x <- m[, c(2,3,6)]; x

#extract 6th to 8th ROW and assign to new variable x
x <- m[6:8, ]; x

#extract elements from [row 2, column 2] to [row 6, column 9] and assign to new variable x
x <- m[2:6, 2:9]; x
```

## Challenge 3
```{r challenge 3}
#construct four dimensional, 400 element array (5x5x4x4) consisting of numbers descending from 400 to 1
a <- array(data = 400:1, dim = c(5,5,4,4)); a #outputs 16 matrices

a[1,1,1,2]
#returns value 300; comes from row 1 and column 1 (1,1) of matrix (1,2)

a[2,3,2,]
# returns [1] 364 264 164 64
# elements from row 2 and column 3 (2,3) of matrices (2, 1:4)

a[1:5, 1:5, 3, 3]
#returns elements from rows and columns 1:5, aka all elements, of (3,3) matrix
```

## Challenge 4
``` {r challenge 4}
# list for strepsirrhines
Lemuroidea <- c("Cheirogaleidae", "Lepilemuridae", "Indriidae", "Lemuridae", "Daubentoniidae")
Lorisoidea <- c("Lorisidae", "Galagidae")
Strepsirhini <- list(Lemuroidea, Lorisoidea) #nesting within streps
  names(Strepsirhini) <- list("Lemuroidea", "Lorisoidea") #naming the list
Strepsirhini #final output

# making lists for haplorhines
Platyrrhini <- c("Cebidae", "Atelidae", "Pitheciidae")
Catarrhini <- c("Cercopithecidae", "Hylobatidae", "Hominidae")
Tarsioidea <- c("Tarsiidae")

# nesting and namining lists for haplorhines
Anthropoidea <- list(Platyrrhini, Catarrhini)
  names(Anthropoidea) <- list("Platyrrhini", "Catarrhini")
Haplorhini <- list(Anthropoidea, Tarsioidea)
  names(Haplorhini) <- list("Anthropoidea", "Tarsioidea")
Haplorhini

# making final list for all primates
Primates <- list(Haplorhini, Strepsirhini)
  names(Primates) <- list("Haplorhini", "Strepsirhini")
Primates
```

## Challenge 5
```{r challenge 5}
m3 <- matrix(data=c(3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2), nrow=5, ncol=3, byrow=TRUE); m3
m3_df <- data.frame(m3); m3_df #coerce matrix to data frame
m3_df$X2 <- as.logical(m3_df$X2) #coerce second column to be logical-valued
m3_df$X3 <- as.factor(m3_df$X3) #coerce third column to be factor-valued
```

<br>
### MODULE 4 NOTES [working with other data structures]

## Matrices and arrays
```{r}
m <- matrix(data = c(1, 2, 3, 4), nrow=2, ncol=2); m
m <- matrix(data=c(1, 2, 3, 4, 5, 6), nrow=2, ncol=3, byrow=FALSE); m

v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2); m1 #bind by row
m2 <- cbind(v1, v2); m2 #bind by column

class(m1)
dim(m1) #tells us there's 2 rows and 4 columns
class(m2)
dim(m2) #tells us there's 4 rows and 2 columns
colnames(m2)
rownames(m2)
str(m2) #summarizes aspects of data structure

a <- array(data = 1:90, dim = c(5,6,3)); a #5 rows, 6 columns, broken up into 3 sets
```

## Subsetting 
```{r}
v <- 1:100; v #creates vector containing all integers from 1 to 100
v [1:15] #select first 15 elements
v[c(2, 4, 6, 8, 10)] #subset specific values
v <- 101:200 #create vector containing all integers from 101-200
v[seq(from = 1, to = 100, by = 2)] #create vector starting at 1, ending at 100, count by 2
```

## Overwriting
``` {r overwriting}
m <- matrix(data = 1:80, nrow =8, ncol = 10, byrow = FALSE); m #pasting from challenge 2 because i don't think r knew what matrix i was calling when i tried to knit, "subscript out of bounds error"
m[7, 1] <- 564; m
# replaces value (7) in row 7, column 1 with value 564
m[, 8] <- 2; m 
# replaces all values in column 8 with value 2
m[2:5, 4:8] <- 1; m 
# replaces all values in rows 2-5, columns 4-8 with value 1
m[2:5, 4:8] <- c(20, 19, 18, 17); m 
# replaces values in rows 2:5, columns 4:8 with 20, 19, 18, 17, respectively
m[2:5, 4:8] <- matrix(data = c(20:1), nrow = 4, ncol = 5, byrow = TRUE); m 
# same as previous but replaced with descending values 20 to 1
# byrow = TRUE fills matrices by row, default is byrow = TRUE which fills matrices by column
m[, 8] <- c("a", "b"); m 
#replaces values in column 8 with alternating "a" and "b" characters
```

## Lists  
``` {r lists}
s <- c("this", "is", "a", "vector", "of", "strings") #making a vector of strings
m <- matrix(data=1:40, nrow = 5, ncol = 8) #makes matrix w/ values 1-40, split into 5 rows and 8 columns
b <- FALSE #logical statement
l <- list(s, m, b); l #combines s, m, b into a list
l[[2]] #references m aka second element in list
l[[2]][2,6] #same as before but isolates entry from row 2, column 6 of matrix
l[1:2] #references s, m from list
l[c(1,3)] #only taking s, b aka first and third elements
str(l) #provides details about 3 elements in list
names(l) <- c("string", "matrix", "logical"); names(l) #names elements in list
l$string #specifically calls element named string in list l
l$matrix[3,5] #calls entry in row 3, column 5 of element named matrix in list l
```

## Data frames
```{r data frames}
df <- data.frame(firstName = c("Nitin", "Silvy", "Ingrid", "Claire"), major = c("Biology (CM)",
    "Anthropology (Bio)", "Anthropology (Medical)", "Biology (EEB)"), sex = c("M",
    "F", "F", "F"), yearInProgram = c(1, 2, 2, 3))
df
df <- read.csv("~/Documents/GitHub/AN588-Week-2/randompeople.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
#reading preexisting csv file into R
df[, c(1, 3, 4, 11, 12)] #only printing selected columns
str(df) #provides details about selected elements/columns
df[,3] #only showing details about column 3 aka name.first
str(df[,3])
df$name.last #use $ as another way to select certain data frames, can also use brackets
str(df$name.last)
df[3]
str(df[3])
df["name.last"]
str(df["name.last"]) #returns data frame of data for column name.last

#add additional rows or columns to data frame using rbind() and cbind()
df <- cbind(df, id = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16,
    17, 18, 19, 20))
df <- cbind(df, school = c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College",
    "MIT", "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU",
    "Harvard", "MIT", "Harvard", "BU", "UMASS Boston", "Harvard"))
#can also do the same thing by using df$school <- c(...) but school will be added as character vector instead of factor
df
```

## Filtering 
``` {r filtering}
new_df <- df[df$school == "BU", ]; new_df #subset data frame and only show rows where variable school is BU
new_df <- df[df$school == "BU", c("name.last", "name.first", "school")]; new_df #can further subset and only show specific columns
```

## Factors, conversion and coercion, and special data values
``` {r factors and etc, eval = FALSE}
as.character() #convert factor to character data
as.factor() #convert character to factor data
na.rm = TRUE #exclude missing data from calculations
```

<br>
### MODULE 5 NOTES [getting data into R]

## Preliminaries 2
```{r prelim 2, eval = FALSE}
library(readr, curl, readxl, rdrop2, repmis)
#install packages google drive and googlesheets4
#error installing XLConnect, ignore
```

## Loading files
for reading .csv or .txt files from remote servers like dropbox or github
```{r load files, eval = FALSE}
library(readr, curl)

#for .csv file, using {curl} package
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)

#for .txt file, using {curl} package
Sys.setenv(VROOM_CONNECTION_SIZE = 131072 * 2)
f <- "https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.csv"
d <- read_csv(f, col_names = TRUE)
head(d)

#importing data from remote server using {readr}
Sys.setenv(VROOM_CONNECTION_SIZE = 131072 * 2)
library(readr)
f <- "https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.csv"
d <- read_csv(f, col_names = TRUE); head(d)
f <- "https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.txt"
d <- read_tsv(f, col_names = TRUE); head(d)
```